% $Id: conclusion.tex 
% !TEX root = main.tex

%%
\section{Background}
\label{sec:background}

In \ac{RL}, intelligent agents learn to map environmental situations (environment states) to actions that maximize a numerical reward signal received from the environment over the long term~\cite{sutton18}. \ac{RL} problems are commonly formulated as Markov Decision Processes (MDPs), defined by the tuple $\mathcal{M} = \langle \mathcal{S}, \mathcal{A}, P, R, \gamma \rangle$, where:
$\mathcal{S}$ is the state space, comprising all relevant states of the environment;
$\mathcal{A}$ is the action space, i.e., the set of all actions the agent can perform to affect the environment;
$P(s_{t+1} \mid s_t, a_t)$ is the transition probability from state {\color{purple}$s_t$} to {\color{purple}$s_{t+1}$} under action {\color{purple}$a_t$};
$R(s_t, a_t)$ is the reward function, providing the numerical signal ($r$) that encodes the positive or negative impact of taking action {\color{purple}$a_t$} in state {\color{purple}$s_t$} at each execution step $t$; and $\gamma \in [0, 1]$ is the discount factor, which determines the importance of future rewards.

In stationary settings, $P$ and $R$ remain fixed over time, and the agent’s objective is to learn a policy $\pi: \mathcal{S} \to \mathcal{A}$ that maximizes the expected discounted return. However, real-world environments often violate this stationarity assumption: transition dynamics and/or reward functions may shift due to evolving conditions. We model such scenarios as \emph{non-stationary} MDPs, represented by a sequence
\[
\{\mathcal{M}_t\}_{t=1}^\infty
\quad\text{with}\quad
\mathcal{M}_t = \bigl\langle \mathcal{S}, \mathcal{A}_t, P_t, R_t, \gamma \bigr\rangle,
\]
where $P_t$ and/or $R_t$ change at unknown time steps $t$ (``concept drifts''). In this work, we further consider the novel scenario where the agent's action space $\mathcal{A}$ may also change over time. Detecting and adapting to these drifts is central to \ac{CRL}~\cite{khetarpal2022continualreinforcementlearningreview,abel2023definitioncontinualreinforcementlearning}, which treats learning as an ongoing process across a continually changing sequence of tasks.

Q-learning~\cite{watkins92} is a widely used model-free approach in \ac{RL}. The long-term quality of an action performed at a given state is computed iteratively in a series of steps and is represented by a Q value,
$\mathit{Q(s,a)}$.
Formally, each execution step $t$ captures information from the environment and maps it to a state
{\color{purple}$s_t$} $\in \mathcal{S}$ in its state space. It then selects an
action {\color{purple}$a_t$} $\in \mathcal{A}$ from its action space and executes it. The agent
receives a reward {\color{Bittersweet}$r_t$} from the environment when it moves to the next state
$s_{t+1} \in \mathcal{S}$. The reward is used to update the optimality of performing the
action {\color{Mulberry}$a_t$} at state {\color{purple}$s_t$}. The agent's goal is to learn
a policy (\i.e. the best-fit action for each state) that maximizes the reward of the
long-run behavior. The learning rate \lrate{\alpha} determines how much new experiences 
overwrite previously learned experiences, and the discount factor {\color{RoyalBlue} $\gamma$} 
determines how much future rewards are discounted so that agents prioritize immediate actions and 
can plan the best long-term actions. At each time step $t$, the Q value of an action 
{\color{purple}$a_{t+1}$} taken in state {\color{purple}$s_{t+1}$}, $Q(s_{t+1}, a_{t+1})$, is updated by 
the Bellman learning equation as follows:

\vspace{1em}
\input{equations/bellman}

While Q-learning converges under stationary conditions (with appropriate decay of \lrate{\alpha}), it can struggle when $P_t$ or $R_t$ change over time. Our work builds on this foundation by incorporating online concept-drift detection and adaptive updates, enabling the agent to remain effective in non-stationary MDPs. We focus particularly on shifts in $R_t$, which may also induce changes in $\mathcal{A}_t$.


\endinput

